---
title: "Lab book: PCA outputs"
output: radix::radix_article
geometry: margin=1in, includeheadfoot
fig_captions: true
fontsize: 3pt
header-includes:
   - \usepackage{longtable}
   - \usepackage{fancyhdr}
   - \fancyfoot[LE,RO]{\thepage}
author:
  - name: Liam Brierley
    affiliation: University of Liverpool
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
---
\pagenumbering{arabic}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, error=TRUE)
```

**PCAs based on genome composition for `r cov_pca_df %>% nrow` coronavirus spike protein sequences. PCAs are colour-coded and ellipses drawn based on different outcome variables, though underlying PCA for each bias type is the same. Mouseover gives outcome variable and virus name.**

## Dinucleotide bias // genus

```{r plot_biplot_dinucs_genus, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(dinucs_pca, "genus")
```

_Fairly clear clustering and separation of genera!_

## Dinucleotide bias // human infection capability

```{r plot_biplot_dinucs_human, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(dinucs_pca, "h_human")
```

_Very tight clusters, especially of SARS-CoV and SARS-Cov-2. MERS, SARS, SARS-CoV-2 actually as distinct from each other as they are from other human CoVs. And many close animal viruses..._

## Codon bias (RSCU) // genus

```{r plot_biplot_codons_genus, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(codons_pca, "genus")
```

_Separation seems driven almost entirely by stop codon use, alphas preferring TGA, betas and gammas preferring TAA, deltas somewhere in between._

## Codon bias (RSCU) // human infection capability

```{r plot_biplot_codons_human, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(codons_pca, "h_human")
```

_Epidemic coronaviruses are strongly separated from each other, but not too separated from other human viruses again. Virtually all the human viruses prefer TAA, except HCoV-HKU1 doesn't seem fussy._

## Amino acid bias // genus

```{r plot_biplot_aa_genus, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(aa_pca, "genus")
```

_Clusters, but not as clear a separation here._

## Amino acid bias // human infection capability

```{r plot_biplot_aa_human, results="asis", layout="l-body-outset", fig.height=10.0, fig.width=10.0}
genome_biplot(aa_pca, "h_human")
```

_Surprisingly, the epidemic human coronaviruses are well separated from the endemic human coronaviruses here!_